---
title: "Linnaeus 2.0"
author: "I. Bartomeus"
output: html_document
editor_options: 
  chunk_output_type: console
---

Note: Do not knit... may produce terribly long output (if works at all).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, message=FALSE, warning=FALSE}
library(reshape2)
library(taxize)
library(traitbaser)
library(originr)
```

This script cleans raw data to explore several questions comparing historical and modern plant surveys.

First we load the data:

Historical data:
```{r load data, message=FALSE, warning=FALSE}
#historical 
lin <- read.csv("raw_data/linneaus.csv", h = T)
head(lin)

#todo: 
#uncap transects
unique(lin$Herbatio)
levels(lin$Herbatio) <- c("GamlaUppsala", "Danmark", "Gottsunda", "Haga",       
                          "Husby", "Jumkil", "Ultuna", "Vaksala")
#check genus and species with taxize...
lin$gen_sp <- paste(lin$Genus, lin$species)
#SKIP from here----- (used last time with an un-updated func)
#I put it in a for loop to avoid time-outs 537
clean_sp <- data.frame(species = NA, matched_names = NA, synonym_names = NA, final_names = NA)
for(i in c(100,200,300,400,500,600)){
  temp <- cleanSpecies(lin$gen_sp[(i-99):i]) #maybe can be repeated with other database. now db == Itis
  clean_sp <- rbind(clean_sp, temp)
}
clean_sp <- clean_sp[-1,]

#save output as is looong to test.
write.csv(clean_sp, "data/clean_linne_sp.csv", row.names = F)
#To HERE-----

#load it when needed
clean_sp <- read.csv("data/clean_linne_sp.csv", h = T)
colnames(clean_sp) <- c("gen_sp", "lin_matched_name", "lin_synonym", "lin_final")

clean_sp[which(is.na(clean_sp$lin_matched_name)),]
clean_sp[-grep(pattern = " ", clean_sp$lin_matched_name),]
clean_sp <- clean_sp[-which(is.na(clean_sp$gen_sp)),]
head(clean_sp)
#to fix a few by Joe

str(lin)
str(clean_sp)
clean_sp2 <- clean_sp[-which(duplicated(clean_sp$gen_sp) == TRUE),]
lin2 <- merge(lin, clean_sp2, all.x = TRUE)
dim(clean_sp2)
dim(lin)
which(!clean_sp$gen_sp %in% lin$gen_sp)
which(!lin$gen_sp %in% clean_sp$gen_sp)
sort(clean_sp$gen_sp) == sort(lin$gen_sp)
dim(lin2) #check!
lin <- lin2

#check other stuff
unique(lin$notes) #great, no notes, no worries
sort(table(lin$gen_sp)) #unique species per transect. Discuss why.
#two repeated:
lin[which(lin$gen_sp == "Lycopodium annotinum"),] #curious
lin[which(lin$gen_sp == " "),] #check why!
#check also with synonims!

#categorize microhabitats
unique(lin$Micro.habitat) #I think catogeries can be made.
#Pratum (Praedium?)
#Palus
#Rupes
#Silva /Pascua
#Agri
#Ripa /fluvius
#...

#Add presence
lin$Presence_lin <- 1

lin <- lin[,-which(colnames(lin) == "notes")]
```

Traits data

```{r load traits, message=FALSE, warning=FALSE}
#species data
traits0 <- read.csv("raw_data/species_level.csv", h = T)
head(traits0)
str(traits0)
#select traits
colnames(traits0)
traits <- subset(traits0, select = c("Species",
                                    "Genus",
                                    "Family",
                                    "open..O..vs.tubular..T.",
                                    "Flower.s.colour",
                                    "P1", #anual perennial
                                    #"L", #light: ignore 
                                    #"F", #moisture: ignore
                                    #"R", #pH
                                    "N" #nitrogen
                                    #, "S" #salt
                                    ))
colnames(traits)[4:7] <- c("Flower_morphology", 
                           "Flower_colour",
                           "annual_perennial", "Nitrogen_use")

#remove explanations at bottom
traits <- traits[1:498,]

#create gen_sp
traits$gen_sp <- paste(traits$Genus, traits$Species)
#remove double spaces and L. Taxize?

#SKIP from HERE-------
#I put it in a for loop to avoid time-outs ~500
clean_sp <- data.frame(species = NA, matched_names = NA, synonym_names = NA, final_names = NA)
for(i in c(100,200,300,400,500)){
  temp <- cleanSpecies(species = traits$gen_sp[(i-99):i]) #maybe can be repeated with other database. now db == Itis
  clean_sp <- rbind(clean_sp, temp)
}
clean_sp <- clean_sp[-1,]
#save output as is looong to test.
write.csv(clean_sp, "data/clean_trait_sp.csv", row.names = F)
#to HERE-----

#load it when needed
clean_trait_sp <- read.csv("data/clean_trait_sp.csv", h = T)
colnames(clean_trait_sp) <- c("gen_sp", "trait_matched_name", "trait_synonym", "trait_final")

#check stuff
clean_trait_sp[which(is.na(clean_trait_sp$trait_matched_name)),]
clean_trait_sp[-grep(pattern = " ", clean_trait_sp$trait_matched_name),]
clean_trait_sp <- clean_trait_sp[-which(is.na(clean_trait_sp$gen_sp)),]
head(clean_trait_sp)
#to fix a few by Joe

#dup
clean_trait_sp[which(duplicated(clean_trait_sp$gen_sp) == TRUE),]
clean_trait_sp[which(duplicated(clean_trait_sp$trait_matched_name) == TRUE),]
clean_trait_sp2 <- clean_trait_sp[-which(duplicated(clean_trait_sp$gen_sp) == TRUE),]
traits2 <- merge(traits, clean_trait_sp2, all.x = TRUE)
dim(traits)
dim(traits2) 

#We may want to check for mating system in treeofsex database.
sex <- read.csv(file = "raw_data/tree_of_sex.csv")
sex$gen_sp <- paste(sex$Genus, sex$species)
table(sex$Selfing..self.incompatible.self.compatible.) #
s <- subset(sex, Selfing..self.incompatible.self.compatible. != "")
s2 <- s[,c("gen_sp", "Selfing..self.incompatible.self.compatible.")]
#why species repated??
length(unique(s2$gen_sp)) #406!
colnames(s2) <- c("gen_sp", "compatibility")
#merge when names match!
which(!s2$gen_sp %in% traits2$trait_matched_name)
traits3 <- merge(traits2, s2, by.x = "trait_matched_name", 
                 by.y = "gen_sp", all.x = TRUE) 
head(traits3)
traits3[-which(is.na(traits3$compatibility)),] #only 8 matches :(
#See for how many Species we are missing traits and add to trait list.

#Need to add origin for the new ones based in fe(){originr}
#traits$origin <- NA
#for(i in 1:length(traits$gen_sp)){
#    res <- flora_europaea(traits$gen_sp[i])
#    traits$origin[i] <- ifelse("Sweden" %in% res$native, "native",
#                          ifelse("Sweden" %in% res$exotic, "exotic",
#                                 "not_in_fe"))
#  }
#Do with final list of plants!
```
Re-surveys

```{r load surveys, message=FALSE, warning=FALSE}

#trails data
trails <- read.csv("raw_data/TrailData.csv", h = T, skip = 1)
head(trails)

#some work to clean this up.
#0) remove totals
colnames(trails)
trails <- trails[,1:36]

#1) remove month names
#colnames(trails) <- gsub(pattern = "May", replacement = "", colnames(trails))
#colnames(trails) <- gsub(pattern = "June", replacement = "", colnames(trails))
#colnames(trails) <- gsub(pattern = "July", replacement = "", colnames(trails))
#colnames(trails) <- gsub(pattern = "August", replacement = "", colnames(trails))

#reshape is not working as I expect... when columns has same name, summing manually:
trails$Danmark <- trails$MayDanmark + trails$JuneDanmark + trails$JulyDanmark + trails$AugustDanmark
trails$GamlaUppsala <- trails$MayGamlaUppsala + trails$JuneGamlaUppsala + trails$JulyGamlaUppsala + trails$AugustGamlaUppsala
trails$Gottsunda <- trails$MayGottsunda + trails$JuneGottsunda + trails$JulyGottsunda + trails$AugustGottsunda
trails$Haga <- trails$MayHaga + trails$JuneHaga + trails$JulyHaga + trails$AugustHaga
trails$Husby <- trails$MayHusby + trails$JuneHusby + trails$JulyHusby + trails$AugustHusby
trails$Jumkil <- trails$MayJumkil + trails$JuneJumkil + trails$JulyJumkil + trails$AugustJumkil
trails$Ultuna <- trails$MayUltuna + trails$JuneUltuna + trails$JulyUltuna + trails$AugustUltuna
trails$Vaksala <- trails$MayVaksala + trails$JuneVaksala + trails$JulyVaksala + trails$AugustVaksala

#2) reshape it
head(trails)
trails_melted <- melt(data = trails,
                      measure.vars = c("Danmark", "GamlaUppsala", "Gottsunda", "Haga", "Husby", "Jumkil", "Ultuna", "Vaksala"),
                      id.vars = c("Family", "Genus", "Species", "Code"),
                      variable.name = "Herbatio",
                      value.name = "Presence")
head(trails_melted)
unique(trails_melted$Herbatio)
subset(trails, Species == "millefolium L.")
subset(trails_melted, Species == "millefolium L.")
trails2 <- dcast(data = trails_melted, 
                 formula = Genus + Species + Family + Herbatio ~ "Presence_us",
                 fun.aggregate = sum,
                 value.var = "Presence") 
head(trails2)
unique(trails2$Presence_us)
#presence can be converted to binary
trails2$gen_sp <- paste(trails2$Genus, trails2$Species)
unique(trails2$gen_sp)
str(trails2)
spec <- unique(trails2$gen_sp)
length(spec)

#SKIP from HERE-------
#I put it in a for loop to avoid time-outs ~500
clean_sp <- data.frame(species = NA, matched_names = NA, synonym_names = NA, final_names = NA)
for(i in c(100,200,300,400)){
  temp <- cleanSpecies(species = spec[(i-99):i]) #maybe can be repeated with other database. now db == Itis
  clean_sp <- rbind(clean_sp, temp)
}
clean_sp <- clean_sp[-1,]

#save output as is looong to test.
write.csv(clean_sp, "data/clean_trails_sp.csv", row.names = F)
#load it when needed
#to HERE-----
clean_trails_sp <- read.csv("data/clean_trails_sp.csv", h = T)
colnames(clean_trails_sp) <- c("gen_sp", "trails_matched_name", "trails_synonym", "trait_final")

#check stuff
clean_trails_sp[which(is.na(clean_trails_sp$trails_matched_name)),]
clean_trails_sp <- clean_trails_sp[-which(is.na(clean_trails_sp$gen_sp)),]
head(clean_trails_sp)
#to fix a few by Joe

#dup
clean_trails_sp[which(duplicated(clean_trails_sp$gen_sp) == TRUE),]
clean_trails_sp[which(duplicated(clean_trails_sp$trait_matched_name) == TRUE),]
clean_trails_sp2 <- clean_trails_sp[-which(duplicated(clean_trails_sp$gen_sp) == TRUE),]

trails3 <- merge(trails2, clean_trails_sp, all.x = TRUE)
dim(trails3)
dim(trails2) #check!

```

We can also use the 'urban_rural.csv' comparision, but I am skipiing it for now. This data is agregated to mean, but in notebook there is more resolution.

Now let's combine the three files

```{r}
# I want a sp, herbatio, habitat, lineus, us, traits spreadsheet.
head(lin,10) #merge by Herbatio and lin_matched_name 
head(trails3, 10) #merge by trails_matched name and Herbatio

# merge by 
lin$link <- paste(lin$lin_matched_name, lin$Herbatio)
trails3$link <- paste(trails3$trails_matched_name, trails3$Herbatio)
colnames(lin)[2] <- "Herbatio_lin"
colnames(trails3)[5] <- "Herbatio_us"
colnames(lin)[1] <- "gen_sp_lin"
colnames(trails3)[1] <- "gen_sp_us"
temp <- merge(lin, trails3, by = "link", all = TRUE)
head(temp) #NEED TO CHECK. LOOKS OK, but NOT MANUEL CHECKING DONE YET
temp$Presence_lin[is.na(temp$Presence_lin)] <- 0
temp$Presence_us[is.na(temp$Presence_us)] <- 0

subset(temp, Presence_us == 0)

head(traits3)
colnames(traits3)[2] <- "gen_sp_traits"
#check next line, assumes good matching...
temp$final_gen_sp <- ifelse(is.na(temp$trails_matched_name), temp$lin_matched_name, temp$trails_matched_name) #unfactor first!
dat <- merge(temp, traits3, by.x = "final_gen_sp", by.y = "trait_matched_name", all.x = TRUE) 
head(dat)

#now we can add origin on here
temp <- unique(dat$final_gen_sp)
temp$origin <- NA
for(i in 1:length(temp$final_gen_sp)){
    res <- flora_europaea(temp$final_gen_sp[i])
    temp$origin[i] <- ifelse("Sweden" %in% res$native, "native",
                          ifelse("Sweden" %in% res$exotic, "exotic",
                                 "not_in_fe"))
  }

```

Save data

```{r}
write.csv(dat, "data/used_data.csv")
```








