---
title: "Linnaeus 2.0"
author: "I. Bartomeus"
output: html_document
editor_options: 
  chunk_output_type: console
---

Note: Do not knit... may produce terribly long output (if works at all).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, message=FALSE, warning=FALSE}
library(reshape2)
library(taxize)
library(traitbaser)
library(originr)
```

This script cleans raw data to explore several questions comparing historical and modern plant surveys.

First we load the data:

Historical data:
```{r load data, message=FALSE, warning=FALSE}
#historical 
lin <- read.csv("raw_data/linneaus.csv", h = T)
head(lin)

### 1. removed entries in linneaus.csv data that had no species
lin <- lin[c(1:294, 296:367, 369:474, 476:495, 497:538),]

#todo: 
#uncap transects
unique(lin$Herbatio)
levels(lin$Herbatio) <- c("GamlaUppsala", "Danmark", "Gottsunda", "Haga",       
                          "Husby", "Jumkil", "Ultuna", "Vaksala")
#check genus and species with taxize...
lin$gen_sp <- paste(lin$Genus, lin$species)
#SKIP from here----- (used last time with an un-updated func)
#I put it in a for loop to avoid time-outs 537
#clean_sp <- data.frame(species = NA, matched_names = NA, synonym_names = NA, final_names = NA)
#for(i in c(100,200,300,400,500,600)){
 # temp <- cleanSpecies(lin$gen_sp[(i-99):i]) #maybe can be repeated with other database. now db == Itis
#  clean_sp <- rbind(clean_sp, temp)
#}
#clean_sp <- clean_sp[-1,]

#save output as is looong to test.
#write.csv(clean_sp, "data/clean_linne_sp.csv", row.names = F)
#load it when needed
#To HERE-----

#Joe cleaning (done once and saved):
clean_linne_sp <- read.csv("data/clean_linne_sp.csv", h = T, stringsAsFactors = FALSE)
head(clean_linne_sp)
## manually change species names using gsub
clean_linne_sp[which(clean_linne_sp$species == "Arbutus Uva urfi"), "synonym_names"] <- "Arctostaphylos uva-ursi"
clean_linne_sp[which(clean_linne_sp$species == "Cerefus Padus"), "synonym_names"] <- "Prunus padus"
## perhaps remove this species as it is a fungus
clean_linne_sp[which(clean_linne_sp$species == "Elvela Mitra"), "synonym_names"] <- "Gyromitra infula"
clean_linne_sp[which(clean_linne_sp$species == "Ophrys Nidus"), "synonym_names"] <- "Neottia nidus-avis"
clean_linne_sp[which(clean_linne_sp$species == "Ranuculus Ficaria:"), "synonym_names"] <- "Ficaria verna"
## perhaps remove this species as it is a lichen
clean_linne_sp[which(clean_linne_sp$species == "Lichen nivalis:"), "synonym_names"] <- "Lepraria nivalis"
clean_linne_sp[which(clean_linne_sp$species == "Ribes Vua crispa"), "synonym_names"] <- "Ribes uva-crispa"
clean_linne_sp[which(clean_linne_sp$species == "TuffiIago Farfara"), "synonym_names"] <- "Tussilago farfara"
clean_linne_sp[which(clean_linne_sp$species == "Convolvulus Polygonum"), "synonym_names"] <- "Polygonum convolvulus"
clean_linne_sp[which(clean_linne_sp$species == "Crataeus Aira"), "synonym_names"] <- "Sorbus aria"
clean_linne_sp[which(clean_linne_sp$species == "Buphtalmo tinctorio"), "synonym_names"] <- "Cota tinctoria"
clean_linne_sp[which(clean_linne_sp$species == "Comarum palustre"), "synonym_names"] <- "Potentilla palustris"
clean_linne_sp[which(clean_linne_sp$species == "Ethusa Cynapium"), "synonym_names"] <- "Aethusa cynapium"
## perhaps remove this species as it is a liverwort
clean_linne_sp[which(clean_linne_sp$species == "Jungermannia pusilla"), "synonym_names"] <- "Blasia pusilla"
clean_linne_sp[which(clean_linne_sp$species == "Ophrys Monorchis"), "synonym_names"] <- "Herminium monorchis"
clean_linne_sp[which(clean_linne_sp$species == "Pedicularis SceptCarolin"), "synonym_names"] <- "Pedicularis sceptrum-carolinum"
clean_linne_sp[which(clean_linne_sp$species == "Scabiosam arvensem"), "synonym_names"] <- "Knautia arvensis"
clean_linne_sp[which(clean_linne_sp$species == "Chenopodium Bonus henricus"), "synonym_names"] <- "Blitum bonus-henricus"
## saved to data folder
head(clean_linne_sp)
write.csv(clean_linne_sp, file="data/clean_linne_sp2.csv")

#load the cleaned data:
clean_sp <- read.csv("data/clean_linne_sp2.csv", h = T, row.names = 1)
head(clean_sp)
colnames(clean_sp) <- c("gen_sp", "lin_matched_name", "lin_synonym", "lin_final")

clean_sp[which(is.na(clean_sp$lin_matched_name)),]
clean_sp <- clean_sp[-which(is.na(clean_sp$gen_sp)),]
clean_sp[-grep(pattern = " ", clean_sp$lin_matched_name),]
head(clean_sp)

str(lin)
str(clean_sp)
clean_sp2 <- clean_sp[-which(duplicated(clean_sp$gen_sp) == TRUE),]
lin2 <- merge(lin, clean_sp2, all.x = TRUE)
dim(clean_sp2)
dim(lin)
clean_sp[which(!clean_sp$gen_sp %in% lin$gen_sp),]
which(!lin$gen_sp %in% clean_sp$gen_sp)
dim(lin2) 
lin <- lin2

#check other stuff
unique(lin$notes) #great, no notes, no worries
sort(table(lin$gen_sp)) #unique species per transect. Discuss why.
#two repeated:
lin[which(lin$gen_sp == "Lycopodium annotinum"),] #curious
lin[which(lin$gen_sp == " "),] #check why!
#check also with synonims!

#categorize microhabitats
unique(lin$Micro.habitat) #I think catogeries can be made.

## 2. grouped factors into categories
## create new column by using micro-habitat data
lin$Habitat_groups <- lin$Micro.habitat
## set data to as factor
lin$Habitat_groups <- as.factor(clean_linneaus$Habitat_groups)
## re group factors in new column
levels(lin$Habitat_groups) <- list(
  Rupes=c("1 Rupes Lassbyenses", "10 Rupes Gottsundenses", "3 Rupes"), 
  Silva=c("2 Pascua Sylvatica", "12 Silva Regia", "5 Silva Regia", 
          "1 Sylva Deserta", "4 Saltus Rickombergensis", "5 Nemus Regium", 
          "6 Lucus Norrbyensis"),
  Palus=c("3 Palus Norrbyensis", "2 Palus Jumkilensis"), 
  Pratum=c("4 Pratum Norrbyense", "7 Pratum Gottsundense", 
                    "11 Praedium Wardsattra", "1 Pratum Regium", 
                    "2 Pratumbarbyense", "3 Prata Ekensia & Fallensia", 
                    "3 Pratum Saltpetterángen", "4 Prata", 
                    "8 Praedium Kiattinge", "9 Pratum Nyflense", 
                    "10 Praedium Husby", "5 Prataflogstadensia & Hagensia"),
  Agros=c("2 Agri Upsalieness", "2 Agri Waxalenses", "3 Versurae Agrorum"),
  Fluvius=c("2 Ripae", "2 Ripa fluvii UPSALIENSIS", "3 Fluvius", 
            "6 Rivulus", "6 Fons Sandwiksensis","5 Uliginosa Amni", 
            "2 Piscina"),
  Urbis=c("1Tecta Upsaliensia", "3 Plateae Urbis Upsaliensis", 
          "1 Via Upsaliensis", "1 Tecta Upsaliensia", "4 Campus Polonicus",
          "Tegelhagen", "1 SEPTA"),
  Rusticus=c("9 Praecipitia Gottsundensia", "6 Lucus Norrbyensis", "7 Haga",
             "1 Luthagen", "5 Tórnby", "4 Falla", "5 Lucus Nantunensis",
             "9 Liljeholmen", "7 Ultuna", "5 Wittulsbergensis",
             "4 Tumuli Upsaliensis", "4 Templo Danemarkensi",
             "7 Templi Borfensis", "3 Lupuleta", "8 Bovilia Pastoris",
             "1 Tegelhagen")
  )
head(lin)

#Add presence
lin$Presence_lin <- 1

lin <- lin[,-which(colnames(lin) == "notes")]
```

Traits data

```{r load traits, message=FALSE, warning=FALSE}
#species data
traits0 <- read.csv("raw_data/species_level.csv", h = T)
head(traits0)
str(traits0)
#select traits
colnames(traits0)
traits <- subset(traits0, select = c("Species",
                                    "Genus",
                                    "Family",
                                    "open..O..vs.tubular..T.",
                                    "Flower.s.colour",
                                    "P1", #anual perennial
                                    #"L", #light: ignore 
                                    #"F", #moisture: ignore
                                    #"R", #pH
                                    "N" #nitrogen
                                    #, "S" #salt
                                    ))
colnames(traits)[4:7] <- c("Flower_morphology", 
                           "Flower_colour",
                           "annual_perennial", "Nitrogen_use")

#remove explanations at bottom
traits <- traits[1:498,]

#create gen_sp
traits$gen_sp <- paste(traits$Genus, traits$Species)
#remove double spaces and L. Taxize?

#SKIP from HERE-------
#I put it in a for loop to avoid time-outs ~500
#clean_sp <- data.frame(species = NA, matched_names = NA, synonym_names = NA, final_names = NA)
#for(i in c(100,200,300,400,500)){
 # temp <- cleanSpecies(species = traits$gen_sp[(i-99):i]) #maybe can be repeated with other database. now db == Itis
  #clean_sp <- rbind(clean_sp, temp)
#}
#clean_sp <- clean_sp[-1,]
#save output as is looong to test.
#write.csv(clean_sp, "data/clean_traits_sp.csv", row.names = F)
#to HERE-----

#Joe cleaning (done once and saved):
clean_traits_sp <- read.csv("data/clean_traits_sp.csv", h = T, stringsAsFactors = FALSE)
head(clean_traits_sp)
## manually change species names using gsub
clean_traits_sp[which(clean_traits_sp$species == "Capsella bursa -pastoris"), "synonym_names"] <- "Capsella bursa-pastoris"
## possibly remove these as they have no positive species ID
clean_traits_sp[which(clean_traits_sp$species == "Arabidopsis sp."), "synonym_names"] <- "Arabidopsis"
clean_traits_sp[which(clean_traits_sp$species == "Myosotis spp."), "synonym_names"] <- "Myosotis"
head(clean_traits_sp)
write.csv(clean_traits_sp, file="data/clean_traits_sp2.csv")

#load it when needed
clean_trait_sp <- read.csv("data/clean_traits_sp2.csv", h = T, row.names = 1)
colnames(clean_trait_sp) <- c("gen_sp", "trait_matched_name", "trait_synonym", "trait_final")
head(clean_trait_sp)

#check stuff
clean_trait_sp[which(is.na(clean_trait_sp$trait_synonym)),]
clean_trait_sp[-grep(pattern = " ", clean_trait_sp$trait_synonym),]
clean_trait_sp <- clean_trait_sp[-which(is.na(clean_trait_sp$gen_sp)),]
head(clean_trait_sp)

#dup
clean_trait_sp[which(duplicated(clean_trait_sp$gen_sp) == TRUE),]
clean_trait_sp[which(duplicated(clean_trait_sp$trait_synonym) == TRUE),]
clean_trait_sp2 <- clean_trait_sp[-which(duplicated(clean_trait_sp$gen_sp) == TRUE),]
traits2 <- merge(traits, clean_trait_sp2, all.x = TRUE)
dim(traits)
dim(traits2) 

#We may want to check for mating system in treeofsex database.
sex <- read.csv(file = "raw_data/tree_of_sex.csv")
sex$gen_sp <- paste(sex$Genus, sex$species)
table(sex$Selfing..self.incompatible.self.compatible.) #
s <- subset(sex, Selfing..self.incompatible.self.compatible. != "")
s2 <- s[,c("gen_sp", "Selfing..self.incompatible.self.compatible.")]
#why species repated??
length(unique(s2$gen_sp)) #406!
colnames(s2) <- c("gen_sp", "compatibility")
#merge when names match!
which(!s2$gen_sp %in% traits2$trait_matched_name)
traits3 <- merge(traits2, s2, by.x = "trait_matched_name", 
                 by.y = "gen_sp", all.x = TRUE) 
head(traits3)
traits3[-which(is.na(traits3$compatibility)),] #only 8 matches :(
#See for how many Species we are missing traits and add to trait list.
#Maybe is a match names issue? 
#MOre work can be done here.


#Need to add origin for the new ones based in fe(){originr}
#traits$origin <- NA
#for(i in 1:length(traits$gen_sp)){
#    res <- flora_europaea(traits$gen_sp[i])
#    traits$origin[i] <- ifelse("Sweden" %in% res$native, "native",
#                          ifelse("Sweden" %in% res$exotic, "exotic",
#                                 "not_in_fe"))
#  }
#Do with final list of plants! See below!
```
Re-surveys

```{r load surveys, message=FALSE, warning=FALSE}

#trails data
trails <- read.csv("raw_data/TrailData.csv", h = T, skip = 1)
head(trails)

#some work to clean this up.
#0) remove totals
colnames(trails)
trails <- trails[,1:36]

#1) remove month names
#colnames(trails) <- gsub(pattern = "May", replacement = "", colnames(trails))
#colnames(trails) <- gsub(pattern = "June", replacement = "", colnames(trails))
#colnames(trails) <- gsub(pattern = "July", replacement = "", colnames(trails))
#colnames(trails) <- gsub(pattern = "August", replacement = "", colnames(trails))

#reshape is not working as I expect... when columns has same name, summing manually:
trails$Danmark <- trails$MayDanmark + trails$JuneDanmark + trails$JulyDanmark + trails$AugustDanmark
trails$GamlaUppsala <- trails$MayGamlaUppsala + trails$JuneGamlaUppsala + trails$JulyGamlaUppsala + trails$AugustGamlaUppsala
trails$Gottsunda <- trails$MayGottsunda + trails$JuneGottsunda + trails$JulyGottsunda + trails$AugustGottsunda
trails$Haga <- trails$MayHaga + trails$JuneHaga + trails$JulyHaga + trails$AugustHaga
trails$Husby <- trails$MayHusby + trails$JuneHusby + trails$JulyHusby + trails$AugustHusby
trails$Jumkil <- trails$MayJumkil + trails$JuneJumkil + trails$JulyJumkil + trails$AugustJumkil
trails$Ultuna <- trails$MayUltuna + trails$JuneUltuna + trails$JulyUltuna + trails$AugustUltuna
trails$Vaksala <- trails$MayVaksala + trails$JuneVaksala + trails$JulyVaksala + trails$AugustVaksala

#2) reshape it
head(trails)
trails_melted <- melt(data = trails,
                      measure.vars = c("Danmark", "GamlaUppsala", "Gottsunda", "Haga", "Husby", "Jumkil", "Ultuna", "Vaksala"),
                      id.vars = c("Family", "Genus", "Species", "Code"),
                      variable.name = "Herbatio",
                      value.name = "Presence")
head(trails_melted)
unique(trails_melted$Herbatio)
subset(trails, Species == "millefolium L.")
subset(trails_melted, Species == "millefolium L.")
trails2 <- dcast(data = trails_melted, 
                 formula = Genus + Species + Family + Herbatio ~ "Presence_us",
                 fun.aggregate = sum,
                 value.var = "Presence") 
head(trails2)
unique(trails2$Presence_us)
#presence can be converted to binary
trails2$gen_sp <- paste(trails2$Genus, trails2$Species)
unique(trails2$gen_sp)
str(trails2)
spec <- unique(trails2$gen_sp)
length(spec)

#SKIP from HERE-------
#I put it in a for loop to avoid time-outs ~500
#clean_sp <- data.frame(species = NA, matched_names = NA, synonym_names = NA, final_names = NA)
#for(i in c(100,200,300,400)){
 # temp <- cleanSpecies(species = spec[(i-99):i]) #maybe can be repeated with other database. now db == Itis
#  clean_sp <- rbind(clean_sp, temp)
#}
#clean_sp <- clean_sp[-1,]
#save output as is looong to test.
#write.csv(clean_sp, "data/clean_trails_sp.csv", row.names = F)
#load it when needed
#to HERE-----

#Joe cleaning (done once and saved):
clean_trails_sp <- read.csv("data/clean_trails_sp.csv", h = T, stringsAsFactors = FALSE)
head(clean_trails_sp)
## manually change species names using gsub
clean_trails_sp[which(clean_trails_sp$species == "Capsella bursa -pastoris"), "synonym_names"] <- "Capsella bursa-pastoris"
## possibly remove these as they have no positive species ID
clean_trails_sp[which(clean_trails_sp$species == "Arabidopsis sp."), "synonym_names"] <- "Arabidopsis"
clean_trails_sp[which(clean_trails_sp$species == "Myosotis spp."), "synonym_names"] <- "Myosotis"
head(clean_trails_sp)
write.csv(clean_trails_sp, file="data/clean_traits_sp2.csv")


#load
clean_trails_sp <- read.csv("data/clean_trails_sp2.csv", h = T, row.names = 1)
head(clean_trails_sp)
dim(clean_trails_sp)
colnames(clean_trails_sp) <- c("gen_sp", "trails_matched_name", "trails_synonym", "trait_final")

#check stuff
clean_trails_sp[which(is.na(clean_trails_sp$trails_matched_name)),]
clean_trails_sp <- clean_trails_sp[-which(is.na(clean_trails_sp$gen_sp)),]
clean_trait_sp[-grep(pattern = " ", clean_trait_sp$trait_matched_name),]
head(clean_trails_sp)

#dup
clean_trails_sp[which(duplicated(clean_trails_sp$gen_sp) == TRUE),]
clean_trails_sp[which(duplicated(clean_trails_sp$trails_synonym) == TRUE),]
clean_trails_sp2 <- clean_trails_sp[-which(duplicated(clean_trails_sp$gen_sp) == TRUE),]

trails3 <- merge(trails2, clean_trails_sp, all.x = TRUE)
dim(trails3)
dim(trails2) 
head(trails3)

```

We can also use the 'urban_rural.csv' comparision, but I am skipiing it for now. This data is agregated to mean, but in notebook there is more resolution.

Now let's combine the three files

```{r}
# I want a sp, herbatio, habitat, lineus, us, traits spreadsheet.
head(lin,10) #merge by Herbatio and lin_synonym 
head(trails3, 10) #merge by trails_synonym name and Herbatio
which(!lin$Herbatio %in% trails3$Herbatio)
which(!trails3$Herbatio %in% lin$Herbatio)
#good
lin[which(!lin$lin_synonym %in% trails3$trails_synonym),] #a lot...
which(!trails3$trails_synonym %in% lin$lin_synonym) #a lot...

# merge by 
lin$link <- paste(lin$lin_synonym, lin$Herbatio)
trails3$link <- paste(trails3$trails_synonym, trails3$Herbatio)
colnames(lin)[2] <- "Herbatio_lin"
colnames(trails3)[5] <- "Herbatio_us"
colnames(lin)[1] <- "gen_sp_lin"
colnames(trails3)[1] <- "gen_sp_us"
lin[which(!lin$link %in% trails3$link),] #some trees!
which(!trails3$link %in% lin$link) #a lot...

temp <- merge(trails3, lin, by = "link", all = TRUE)
head(temp) #NEED TO CHECK. LOOKS OK, but NOT MANUEL CHECKING DONE YET
temp$Presence_lin[is.na(temp$Presence_lin)] <- 0
temp$Presence_us[is.na(temp$Presence_us)] <- 0

head(traits3)
colnames(traits3)[2] <- "gen_sp_traits"
#check matching...
colnames(temp)
temp[which(is.na(temp$trails_synonym)),c(1,16,19)]
#check NA Danmark  etc... 

temp$final_gen_sp <- ifelse(is.na(temp$trails_synonym), as.character(temp$lin_synonym), as.character(temp$trails_synonym)) 
duplicated(traits3$trait_synonym) #5 dup's!!
dat <- merge(temp, traits3, by.x = "final_gen_sp", by.y = "trait_synonym", all.x = TRUE) 
dim(temp)
dim(dat) #!!! FIX
head(dat)

#now we can add origin on here
temp <- data.frame(final_gen_sp = unique(dat$final_gen_sp), origin = NA)
for(i in 1:length(temp$final_gen_sp)){
    res <- flora_europaea(temp$final_gen_sp[i])
    temp$origin[i] <- ifelse("Sweden" %in% res$native, "native",
                          ifelse("Sweden" %in% res$exotic, "exotic",
                                 "not_in_fe"))
}
head(temp, 50) #need some more checking.
#merge
dat2 <- merge(dat, temp, all.x = TRUE)
head(dat2)

#Last clean a bit columns:
dat3 <- dat2[,c(1,3,7,8,12,13,14,20,24:28,31,32)] #PROBABLY I AM MISSING SOME RELEVANT COLUMNS LIKE MICROHABITAT.
head(dat3)

#things can be improved:
#- remove 0's when casting us dataset
#- Refill herbatio's ¿?
```

Save data

```{r}
write.csv(dat3, "data/used_data.csv")
```




