---
title: "analysis"
author: "I. Bartomeus"
date: "10/2/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Analysis:

- Fig 1: Map (with old anc current city extent if posible)

- % of persisting or lost plants per transect. New plants can be exotic, human additions, not recorded by linneaus (CAVEAT). Relate to Vellend et al. richness loss discusion. Fig 2

```{r}
head(dat)


```


- Overall persistance in number of species, showing proportion of each group belonging to anual / tubular / color / N (and maybe self-comp) Fig 3 : Chi Sqr?

- Potential Fig 4: Porp of losers per habitat. Are some habitats more affected than others? Maybe CWM plots? Relate to consequences for pollinators.

- Comparision Urban Rural: Ordination , but also % exotic and % of species persisting (comparing with linne) Fig 5.

-extra: phylogenetic loss... is loss random with respect to phylogeny or clustered in some families?

Discusion points:

[ ] Find population change city area change for uppsala, and how it relates to other cities.
[ ] Maybe test for sampling effort (accumulation curves?). Maybe we can't. Plants detected while in flower. 
[ ] Biotic homogenization: Are "new" recorded plants, specially exotic, common to the 8 trails?
