---
title: "analysis"
author: "I. Bartomeus, ..."
date: "10/2/2017"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Analysis to do:

- Fig 1: Map (with old and current city extent if posible). Not done yet.

- % of persisting or lost plants per transect. New plants can be exotic, human additions, OR not recorded by linneaus (CAVEAT). 
Relate to Vellend et al. richness loss discusion. Fig 2

```{r}
dat <- read.csv("data/used_data.csv", h = T, row.names = 1)
head(dat)
temp <- subset(dat, Presence_lin == 1)
head(temp)
temp$Presence_us0 <- ifelse(temp$Presence_us > 0, 1, 0)

temp$loss <- temp$Presence_lin - temp$Presence_us0
temp$loss #almost all loss???
sum(temp$loss) / length(temp$loss) #86% loss! Something is wrong with data processing...¿?
temp[,c(1,6,8,16,17)]


```


- Overall persistance in number of species, showing proportion of each group belonging to anual / tubular / color / N (and maybe self-comp) 
Fig 3 : test with a Chi Sqr?

- Potential Fig 4: Porp of losers per habitat. Are some habitats more affected than others? Maybe CWM plots? Relate to consequences for pollinators.

- Comparision Urban Rural: Ordination , but also % exotic and % of species persisting (comparing with linne) Fig 5.

-extra: phylogenetic loss... is loss random with respect to phylogeny or clustered in some families?

Discusion points:

[ ] Find population change city area change for uppsala, and how it relates to other cities.
[ ] Maybe test for sampling effort (accumulation curves?). Maybe we can't. Plants detected while in flower. 
[ ] Biotic homogenization: Are "new" recorded plants, specially exotic, common to the 8 trails?
